---
title: "ANLY503 Final Project: Tornado study 1950-2018"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
library(dplyr)
library(tidyr)
library(maps)
library(ggplot2)
library(gganimate)
library(igraph)
library(RColorBrewer)
```

About
===================================== 
Data Source: https://www.spc.noaa.gov/wcm/

This project is aimed to analyze the distribution of tornado occurrence in the US.
By several visualizations, this project would explore details of all tornadoes across the US from 1950 to 2018.

Packages used in this project include:

flexdashboard

dplyr

tidyr

maps

ggplot2

gganimate

igraph

RColorBrewer


Visualizations
===================================== 
Row {data-height=600}
-----------------------------------------------------------------------
### As shown in the choropleth, states in the mid-south are affected by tornadoes more than the others.Texas has the most tornado occurrence in all times


```{r}
df = read.csv('data/1950-2018_all_tornadoes.csv')
df = na.omit(df)

#two functions that transform abbreviations into full names
abbr2state = function(abbr){
  ab    = tolower(c("AL",
             "AK", "AZ", "KS", "UT", "CO", "CT",
             "DE", "FL", "GA", "HI", "ID", "IL",
             "IN", "IA", "AR", "KY", "LA", "ME",
             "MD", "MA", "MI", "MN", "MS", "MO",
             "MT", "NE", "NV", "NH", "NJ", "NM",
             "NY", "NC", "ND", "OH", "OK", "OR",
             "PA", "RI", "SC", "SD", "TN", "TX",
             "CA", "VT", "VA", "WA", "WV", "WI",
             "WY", "DC"))
  st    = c("Alabama",
             "Alaska", "Arizona", "Kansas",
             "Utah", "Colorado", "Connecticut",
             "Delaware", "Florida", "Georgia",
             "Hawaii", "Idaho", "Illinois",
             "Indiana", "Iowa", "Arkansas",
             "Kentucky", "Louisiana", "Maine",
             "Maryland", "Massachusetts", "Michigan",
             "Minnesota", "Mississippi", "Missouri",
             "Montana", "Nebraska", "Nevada",
             "New Hampshire", "New Jersey", "New Mexico",
             "New York", "North Carolina", "North Dakota",
             "Ohio", "Oklahoma", "Oregon",
             "Pennsylvania", "Rhode Island", "South Carolina",
             "South Dakota", "Tennessee", "Texas",
             "California", "Vermont", "Virginia",
             "Washington", "West Virginia", "Wisconsin",
             "Wyoming", "District of Columbia")
  st[match(tolower(abbr), ab)]
}

int2month = function(input){
  int = c(1,2,3,4,5,6,7,8,9,10,11,12)
  month = c('January','February','March','April','May','June','July','August','September','October','November','December')
  
  month[match(input,int)]
}

#transform state and month abbr into full names
df = df %>% mutate(st = tolower(abbr2state(st)))
df = na.omit(df)
df = df %>% mutate(mo = int2month(mo))

#Geospatial data for choropleth
df$region = df$st
states = map_data('state')

#Count tornadoes by year and state
d2 = count(df,yr,region)

#Get all state names and year
all_st = unique(d2$region)
all_yr = unique(d2$yr)

#Fill in 0 for those states that had no tornado in that year
for(i in all_yr){
  cell = d2[d2$yr==i,]
  diff = setdiff(all_st,unique(cell$region))
  for(x in diff){
    d2 = rbind(d2, data.frame(yr=i,region=x,n=0))
  }
}

#Prepare data for choropleth
d3 = left_join(states,d2)

#Plot choropleth with animation
#This could take a while to run
g1 = ggplot(d3,
       mapping = aes(x = long, y = lat,
                    group = group)) + 
  geom_polygon(aes(fill = n)) +
  theme_void() +
  labs(title = 'Number of Tornadoes across the US in Year: {frame_time}',fill = 'number of tornadoes') +
  transition_time(yr)

ani_g1 = animate(g1,fps = 1, height = 375, width = 600)
ani_g1
```

### Tornadoes that originate in Oklahoma, Texas, Alabama, Kansas, and Mississippi are more likely to affect neighbor states


```{r}
#Find tornadoes that also travel to other states
two_st_t = df[df$ns == 2,]

#Get the index for those tornadoes
two_om = unique(two_st_t$om)

#Create new dataframe for source-target network data
network = data.frame(target=character(),source=character())

for(om in two_om){
  cell = two_st_t[two_st_t$om==om,]
  source = cell[1,]$st
  target = cell[nrow(cell),]$st
  network = rbind(network,data.frame(source=source,target=target))
}
network = network %>% filter(source != target)

#Find top 5 states that have tornadoes traveling to other states
source = count(network,source)
source = source[order(-source$n),]
src_5 = head(source,5)

#Count occurance and get the state names
all_counts = count(network,source,target)
top5 = all_counts %>% filter(source %in% src_5$source)

#Plot network
net1 = graph_from_data_frame(d=top5)
V(net1)$size = 10
V(net1)$label.cex = 0.8
V(net1)$label.color = 'black'
V(net1)$frame.color = 'white'
V(net1)$color = brewer.pal(n=8,name = "Set2")
E(net1)$arrow.size = 0.1
E(net1)$width = E(net1)$n/2.5
E(net1)$curved = 0.2
E(net1)$color = 'lightsteelblue1'
plot(net1,layout=layout.circle, 
     main='Top5 tornado origins that affects other states',
     sub=paste('top5 origins:',paste(src_5$source,collapse = ',')))
```


Row {data-height=400}
-----------------------------------------------------------------------

### From this bar chart, it shows that tornadoes occurs mostly in spring times, in April, May, and June


```{r}
#Plot a bar chart showing tornado numbers in each month, with animation on year
mo = count(df,yr,mo) %>% ggplot(aes(x= reorder(mo,-n), y=n, fill = mo)) +
  geom_bar(stat='identity') +
  geom_text(aes(label = n), vjust=-0.2) +
  labs(x = 'month', y = 'number of tornadoes', fill = 'month', title = 'Number of Tornadoes in each Month in Year: {frame_time}') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45)) +
  transition_time(yr) +
  ease_aes('linear')

ani_mo = animate(mo,fps=1,height = 375, width = 600)
ani_mo
```

### Over time, tornado occurrence appears to increase. The number hits its top in 2004, and starts going down


```{r}
#Plot a line graph showing tornado numbers over year
#Regression line included
df_yr = count(df,yr)
yr = df_yr %>% ggplot(aes(x = yr, y = n)) +
  geom_line(color = 'steelblue',size=0.75) +
  geom_point(size = 1) +
  labs(x = 'year', y = 'number of tornadoes', title = 'Number of Tornadoes in each Year 1950-2018') +
  geom_smooth(se=F, color = 'darkgreen') +
  annotate('text',
           x = df_yr[which(df_yr$n == max(df_yr$n)),'yr'],
           y = df_yr[which(df_yr$n == max(df_yr$n)),'n'] + 100,
           label = paste('max value at year',df_yr[which(df_yr$n == max(df_yr$n)),'yr'],':',df_yr[which(df_yr$n == max(df_yr$n)),'n'])) + 
  theme_minimal()

yr
```

### Most of the tornadoes are not very large, as points cluster in the bottom left corner. The magnitudes are mostly smaller than 3


```{r}
#Plot a scatter plot showing tornado size distributions
#Color indicates tornado magnitude, with 5 as maximum, -9 as unknown
mag = df %>% 
  select(mag,len,wid) %>%
  ggplot(aes(x = wid,y=len)) +
  geom_point(aes(color = as.factor(mag))) + 
  scale_color_manual(name='F-scale',
                     values = c('5'='red',
                                '4'='orange',
                                '3'='blue',
                                '2'='yellow',
                                '1'='darkgreen',
                                '0'='lightgreen',
                                '-9'='black'),
                     labels = c(-9,0,1,2,3,4,5)) +
  labs(title='All Tornadoes magnitude',x='width (in yards)',y='length (in miles)') +
  theme_minimal()
  
mag
```







