---
title: "ANLY503 HW5 Wenhao Jiang"
author: "Wenhao Jiang"
date: "2020/10/23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=F,message=F)
library(reticulate)
use_condaenv('r-reticulate',required = T)
```

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
```

```{r}
trc = read.csv('data/transactions.csv')
acc = read.csv('data/accounts_analytical.csv')
```
# Loan amount histogram
```{r}
acc %>%
  ggplot(aes(x=loan_amount)) +
  geom_histogram()
```
# Loan Distribution with status
```{r}
acc %>% subset(!is.na(loan_default)) %>%
  ggplot(aes(x=loan_amount, y=loan_status, color=loan_default)) +
  geom_point() +
  facet_grid(~loan_default)
```

# Loan Distribution if default
```{r}
acc %>% subset(!is.na(loan_default)) %>%
  ggplot(aes(x=loan_amount, y=loan_payment, color=loan_term)) +
  geom_point() +
  facet_grid(~loan_default)
```

1. Loan amount and loan payment with different terms has their own patterns - following one line per term
2. Normally, loan with high amount would not have low payment. However, default loans would have relatively lower payment compared to non-default loans
3. Loan amount distribution is right skewed


# Python representation
```{r, include=F}
py_install('matplotlib')
py_install('seaborn')
py_install('pandas')
```

```{python}
import matplotlib.pyplot as plt
import seaborn as sns
import pandas as pd

df = pd.read_csv('data/accounts_analytical.csv')
sns.relplot(data=df,x='loan_amount',y='loan_payment',col='loan_default',hue='loan_term')
plt.show()
```








